% Public source snapshot of the instructor's assignment-oriented Module 11 slides.
% The overlapping RLHF/DPO survey, local preamble, diagrams, external-library code,
% assignment-specific product configuration, product-current claims, and template
% listing are omitted under sources/README.md. The retained model-card and
% slice-evaluation material documents provenance; publication claims are independently
% checked against primary sources.
\documentclass[xcolor=svgnames, 10pt, handout]{beamer}
\usepackage{amsmath}
\usepackage{amssymb}

\title{Lecture 11.4: Reinforcement Learning for LLM Fine-Tuning}
\subtitle{RLHF, DPO, and the Path to Alignment}
\author{Deep Learning Class}
\date{}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}{Public Snapshot Boundary}
\begin{alertblock}{Provenance, not a publishable asset bundle}
This snapshot preserves the instructor-authored model-card and slice-evaluation
lecture material used as a Chapter 18 seed. The overlapping RLHF/DPO survey, local
preamble, diagram source, external-library implementations, assignment-specific
product configuration, product-current claims, and template listing have been
omitted; their original locations remain marked. Remaining claims are historical
lecture notes and are independently checked before book publication.
\end{alertblock}
\end{frame}

% [Earlier RLHF/DPO survey omitted from this public snapshot: it overlaps the
% independently sanitized 11.4-reinforcement-learning.tex and 11L-llm-alignment.tex
% snapshots. Diagram source, external-library code, assignment-specific product
% configuration, and product-current adoption claims are also omitted under
% sources/README.md.]

\section{Model Cards and Slice-Aware Evaluation}

\begin{frame}{Model Cards: Documentation for Responsible AI}

\textbf{What is a model card?}

\pause

A structured document that reports:
\begin{itemize}
\item \textbf{Intended use:} What problem? Who are users?
\pause
\item \textbf{Data:} Sources, licenses, preprocessing
\pause
\item \textbf{Metrics:} Performance (overall \& by subgroup)
\pause
\item \textbf{Limitations:} Known failure modes, biases
\pause
\item \textbf{Versioning:} Changelog, contact info
\end{itemize}

\pause

\vspace{0.3cm}

\textbf{Analogy:} Model card = ``nutritional label'' for ML models

\end{frame}

\begin{frame}{Why Model Cards Matter}

\textbf{Transparency:} Users understand capabilities and limitations

\pause

\textbf{Responsible deployment:} Prevents misuse by documenting scope

\pause

% [Product-current governance and legal-compliance claim omitted; publication claims
% are checked against current primary sources when needed.]

\textbf{Reproducibility:} Others can replicate or build on your work

\pause

% [Assignment-specific requirement omitted from the public snapshot.]

\end{frame}

\begin{frame}{Model Card: Required Sections}

\textbf{Minimum viable model card:}

\begin{enumerate}
\item \textbf{Model details:} Architecture, size, base model
\pause
\item \textbf{Intended use and limitations:} Clear scope
\pause
\item \textbf{Data:} Sources, dates, licenses
\pause
\item \textbf{Metrics:} Perplexity, toxicity, slice-aware evaluation
\pause
\item \textbf{Training details:} Hyperparameters, hardware
\pause
\item \textbf{Risks and biases:} Known issues
\pause
\item \textbf{Contact:} Maintainer, version
\end{enumerate}

\end{frame}

\begin{frame}{Model Card Template (Omitted)}

% [Assignment-specific model-card template listing omitted from the public snapshot;
% independent provenance was not established and product/version details go stale.]

The original lecture included a simplified assignment template. Its surrounding
field definitions are preserved in the adjacent frames; the listing itself is not
part of this provenance snapshot.

\end{frame}

\begin{frame}{Slice-Aware Evaluation}

\textbf{Don't just report overall metrics!}

\pause

For a metric $M$ (accuracy, perplexity) evaluated on groups $g \in \mathcal{G}$:

\[
\Delta_{\text{parity}}(M) = \max_{g \in \mathcal{G}} M_g - \min_{g \in \mathcal{G}} M_g
\]

\pause

\textbf{Example:} If perplexity is:
\begin{itemize}
\item Domain A: 10.5
\item Domain B: 15.2
\end{itemize}

\pause

$\Delta_{\text{parity}} = 15.2 - 10.5 = 4.7$ (significant gap!)

\pause

\vspace{0.3cm}

\textbf{Best practice:} Report performance \textit{by slice} + overall

\end{frame}

% [Later assignment instructions, executable workflows, product documentation links,
% and assignment-specific deployment guidance omitted from this public snapshot.]

\begin{frame}{Further Reading}

\textbf{Papers:}
\begin{itemize}
\item InstructGPT/RLHF: Ouyang et al. (2022)
\item DPO: Rafailov et al. (2023)
\item Model Cards: Mitchell et al. (2018)
\end{itemize}

% [Product documentation and course recommendations omitted from the public snapshot.]

\end{frame}

\end{document}
